# RStudio and RStudio Projects {#rstudio}

Much of this section is taken from Nick Tierney's excellent book [R Markdown for
Scientists](https://rmd4sci.njtierney.com/rstudio-what-and-why.html).

Here, we will:

  - Look at the structure of an RStudio window,
  - Set some useful default options for RStudio,
  - Create our first RStudio project

## RStudio

R is a programming language, like python, or C. One can write R code in whatever
they feel like: Notepad, VSCode, punch card (theoretically). But, one of the
most popular ways to write R code is using the RStudio IDE. An Integrated
Development Environment (IDE) is just a neat program that helps you write code
more accurately and faster. RStudio has been around for a long time, and has a
lot of R-specific features to help you write solid R code.

You should both R and RStudio installed already. If you open RStudio, you'll be
presented with a window something like this:

``` {r, fig.cap="A screenshot of the RStudio working environment. Image from https://rmd4sci.njtierney.com/rstudio-what-and-why.html."}
image_from_url("https://rmd4sci.njtierney.com/figs/rstudio-screenshot.png")
```

Top left is the source window - this is where your scripts or markdown files
will sit to be edited.

Bottom left is the console, where your code is executed. It also has an
integrated terminal which can be useful for running certain specific tasks.

Top right is the "Viewer", which has a file browser, plot viewer, help window
and a handful of other tabs. If you draw a plot in the console it will appear
here for example.

Bottom right is the "Environment" pane, which will list all the variables you
have loaded currently, the history of your previous commands, and a handful of
other things such as version control, build tools and database connections.

### Exercise 2.1

Open up the RStudio preferences (Tools \> Global Options), and change the layout
of your panes! You can arrange them whatever way you like.

## RStudio Default Options

RStudio by default is very helpful. It will keep items in your environment if
you quit (or crash, usually), and reload them when you re-open RStudio.

However, this can be a bit of a reproducibility nightmare. So, we will turn them
off.

Go to the RStudio preferences (Tools \> Global Options), and set the following:

  - Workspace:
      - Uncheck "Restore .Rdata into workspace upon startup"
      - Save workspace .Rdata on exit: "Never".
  - History:
      - Uncheck "Remove duplicate entries in history"

Some people will suggest unchecking "Always save history (even when not saving
.Rdata)", but I have found personally I rely on history a lot and so not saving
that causes more problems than it's worth. Your mileage may vary.

I'd also recommend changing the colour scheme to a less retina burning white. My
personal favourite is Cobalt currently.

## Rstudio Projects

Now that we have our RStudio set up nicely, it's time to start work on our
project!

**Not so fast!**

We should think about how we will arrange our files. A bit of time now will save
us a lot of pain later. As someone who has experienced a lot of file management
related pain, I cannot stress this enough.

### Never use `setwd()`\*

\*(almost never anyway)

It is the least transportable way to set where R is currently looking for files.
For example, think about this piece of code:

``` {r, eval=F, echo=T}
setwd("/Users/mlyd0001/projects/ResearchToolsWorkshop")
read_csv("data/mtcars.csv")
```

This will set the working directory to this workshop, and then load in a CSV I
have saved in. Works fine for me. But, what if Mitch would like to work from the
same data I'm using? No problem he can just download my materials and run the
script.

    Error in setwd("/Users/mlyd0001/projects/ResearchToolsWorkshop") : cannot change working directory

Oh. Why would that be?

Well, Mitch isn't `mlyd0001`, and so that folder simply doesn't exist on their
computer. So straight away, my collaborator will have to go and edit the script,
which when they send it back to me, will just error on my end!

Perhaps Jenny Bryan feels [a bit too
strongly](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/) about this
concept, but:

> If the first line of your R script is
> `setwd("C:\\Users\\jenny\\path\\that\\only\\I\\have")` I will come into your
> office and SET YOUR COMPUTER ON FIRE 🔥.

And while I won't set your computer on fire, I will edit all your code to remove
it before I send it back to you, and probably passive-aggressively mention it in
my response.

### Project structure

At the end of the day, how you structure your project is up to you (except for
`setwd()`, that isn't up to you). But, we should keep all our files organised in
one, top-level folder.

My personal structure looks something like this:

    project_name
      └── R/
          └── various .R files, usually functions
      └── data/
          └── whatever data files I happen to need
      └── outputs/
          └── figures/
          └── tables/
      └── libraries.R
      └── pipeline.R

If it is, then I will break up `pipeline.R` into multiple files, to produce the
given outputs. But it is critical that the pipeline (or it's subfiles) run from
top to bottom *in a clean R session*.

The clean R session is important, as it will ensure you're not accidentally
relying on anything left over in your Environment (remember the option we
changed earlier about saving the `.Rdata`?) and it means that when your
collaborator runs the file you know it will run from the top.

A handy trick to source all your functions in the pipeline is to use the
following:

``` {r, eval=F, echo=T}
invisible(lapply(list.files("R/", pattern = "*.R", full.names = T), source))
```

Just remember to re-source if you edit and save.

## Projects, with a capital P

So far, we've talked about project structure in an 'informal' sense. RStudio
takes it a step further, with a formal "Project" (the P is capitalised on
purpose).

These projects do a few things:

  - Store project specific settings
  - Project-specific .Rhistory, .Rprofile and a handful of others
  - Add a handy little piece of text on your RStudio icon
  - Store it's own history
  - Set your working directory to the project-specific top level folder.

They're pretty neat.

They also come with their own file, ending in `.Rproj`, which you can open
directly to open the project in RStudio instead of loading through the menu.

*RStudio Projects are specific to RStudio*. If you change IDE or choose not to
use RStudio, they will not do anything. Other IDEs do have similar concepts:
VSCode has "workspaces", PyCharm has "ideas", but they are not cross-compatible.

### Exercise 2.4

Set up an RStudio Project for this workshop!

Go to File \> New Project. Usually you'll be in "New directory", then New
project.

When you name a project, I think it's always worth taking a little bit of time
to think of a nice name. For analysis projects (that aren't software release
based), I tend to just use descriptive names, like "ResearchToolsWorkshop". Good
names will make it easy to find your things years later.

The best names apparently:

  - are short
  - do not use spaces
  - cleverly combine words

## Making sure you're in the right place

While we're working in an RStudio Package, it's unlikely we'll need to change
our working directory. But it can happen sometimes - maybe we are working on a
remote server, or one of our "collaborators" has left in a pesky `setwd()` call
again.

We can ensure we're loading data from the right place using the handy package
`here`.

The job of `here` is simple. The function `here::here()` uses some magic^[not
actually magic, just an algorithm and a specific set of files. See
https://github.com/r-lib/rprojroot for more info.] to determine where our
project root is.

Here is my path:

``` {r, echo=T}
here::here()
```

And if I wanted to load some data, I could use

``` {r, echo=T}
here::here("data/people_who_have_used_setwd.csv")
```

and then no matter how many people have changed working directory on me, `here`
will resolve back to the top-level folder.

The extra nice thing about `here` is that it will work even if we're not in an
RStudio project. The types of files it searches are more than `.Rproj` and so it
will work in *almost* all contexts and IDEs.
