# Avoiding dependency hell with `renv`

Have you ever downloaded someone else's code from a long time ago, and it relies
on packages for a version of R 12 years ago?

Have you then tried to install these old versions of packages only to be met
with an error telling you you can't downgrade that package because 26 other
packages rely on it? (I'm looking at you `rlang`)

This is what is termed *dependency hell*. As always, there is an XKCD for that
(Figure \@ref(fig:python-dependency-hell)). It happens when there is a complex
series of dependencies, often many loops, and occasionally different version
requirements. R is surprisingly OK with this sort of thing, but if you have a
lot of projects (and particularly old ones), then dependency hell can be a real
issue.

``` {r python-dependency-hell, fig.cap="As always, there's an XKCD for that. Image from https://xkcd.com/1987/."}
image_from_url("https://imgs.xkcd.com/comics/python_environment.png")
```

We can avoid dependency hell for our projects by creating *isolated*, *fixed*
environments for our packages. The RStudio team have developed the package
`renv` to help with exactly that.

`renv` effectively creates a local library for your project, and ensures that
libraries are only loaded from that local source, rather than from a single,
global source, as is the default.

This means that you can have different versions of packages installed for
different projects, without any conflicts.

## `renv` basics and quick-start

We start by installing and initialising the project library:

``` {r, renv-install, eval=F, echo=T}
install.packages("renv")
library(renv)
renv::init()
```

You will get a message something like:

    * Initializing project ...
    * Discovering package dependencies ... Done!
    * Copying packages into the cache ... Done!

  - Lockfile written to '\~/projects/ResearchToolsWorkshop/renv.lock'.

Restarting R session...

``` 
```

and there may be a big list of packages that have been detected automatically.

From now on, the basic set up is to install a package using `renv::install()`,
just as we would do normally. If installation succeeds, we call
`renv::snapshot()`, and the versions of our packages are saved into the
*lockfile*.

If we need to revert our packages, or install somewhere else, we can call

``` {r, eval=F, echo=T}
renv::restore()
```

and all the packages will be installed (at the correct versions), and we can be
on our merry way.

## What is a lockfile, anyway?

The lockfile contains all in the information for the packages in the current
environment. It is a JSON file, and human readable. Here's the entry for the
version of `knitr` used in this book:

    "knitr": {
          "Package": "knitr",
          "Version": "1.37",
          "Source": "Repository",
          "Repository": "CRAN",
          "Hash": "a4ec675eb332a33fe7b7fe26f70e1f98",
          "Requirements": [
            "evaluate",
            "highr",
            "stringr",
            "xfun",
            "yaml"
          ]
        }

There's a good amount of information in here. The name of the package, the
version, where it came from (CRAN in this case), as well as the dependencies (or
"requirements").

There's also a "hash", which is the md5sum which will be unique to the zip file
downloaded as part of the installation process. We don't need to worry too much
about it, it's just another error checking step available.

Importantly, because the lockfile is plaintext, it is easily managed in version
control. More on that later...

## Practice time

Inside the RStudio Project we made in the previous chapter, let's initialise a
`renv` environment.

``` {r, echo=T, eval=F}
install.packages("renv")
library(renv)
renv::init()
renv::install("tidyverse")
renv::snapshot()
```
